A tractable model for D-optimal design in infinite dimensional Bayesian inverse problems.

### 2020-07-20 paper.tex
Add tikz image

\subsection{Characterizing D-Optimal Designs}\label{subsub:characterization}
In light of \eqref{eq:mod conditions}, let us denote eigenvalues of
$\fwd \prcov \fwd^*$ by $\lambda_1 \geq \lambda_2 \geq ...$ and its
eigenvectors $\{\ev_i\}_{i=1}^{\infty}$. First, let us present a lemma:

@@

Thus, if we take $\func_j^{*} = \meas_j$ and $C := (\fwd \prcov
\fwd^*)^{-1}$ in Lemma \ref{lemma:sim diag}, we conclude from the
eigenproblem \eqref{eq:mod conditions} that $\obs^*\obs$ has the same
eigenvectors as $\fwd \prcov \fwd^*$, namely
$\{\ev_i\}_{i=1}^{\infty}$ (infinitely many of them with zero
eigenvalue, since $\obs^*\obs$ is finite-rank). Denote the
corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$. By the argument
in the paragraph following \eqref{eq:mod conditions}:
\begin{align}\label{eq:pre manipulations}
  \begin{split}
    I + \sigma^{-2} \prcov^{1/2}  \fwd ^* \obs^* \obs \fwd \prcov^{1/2} &= 
    \prcov^{1/2} ( \prcov^{-1} + \sigma^{-2} \fwd^*\obs^*  \obs \fwd ) \prcov^{1/2}\\
    %
    %
    %
    &=\prcov^{1/2} \fwd^* ( \fwd^{-*} \prcov^{-1}\fwd^{-1} + \sigma^{-2} \obs^* \obs  ) \fwd \prcov^{1/2}\\
    %
    %
    %
    &= \prcov^{1/2} \fwd^* \left ( (\fwd \prcov \fwd^* )^{-1} +\sigma^{-2} \obs^* \obs  \right ) \fwd \prcov^{1/2}.
  \end{split}
\end{align}
By definition of the objective \eqref{eq:objective} and using
\eqref{eq:pre manipulations}:
\begin{align}\label{eq:mod objective}
  \begin{split}
    \tar ( \obs ) &=\frac12 \log \det \left( I + \sigma^{-2}
    \prcov^{1/2} \fwd ^* \obs^* \obs \fwd \prcov^{1/2} \right )\\
    %
    %
    %
    &=\frac12 \log \det
    \left ( (\fwd \prcov \fwd^* )^{-1} + \sigma^{-2}\obs^* \obs  \right ) \fwd \prcov \fwd^* \\ 
    %
    %
    %
    &=\frac12 \log \left (\prod_{i=1}^{\infty} ( \lambda_i^{-1} + \sigma^{-2} \eta_i ) \prod_{i=1}^{\infty} \lambda_i \right ) \\
    %
    %
    %
    &=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right ). 
  \end{split}
\end{align}
Therefore, an optimal design maximizes
\begin{equation}
  \sum_{i=1}^{k}\log (\lambda_i^{-1} + \sigma^{-2} \eta_i),
\end{equation}
with the norm constraints on $\meas_j,j=1,...,m$. We show in the
appedix (Lemma \ref{lemma:free}) we can find $\meas_1,...,\meas_m$
such that $\{\eta_i\}_{i=1}^k$ have any value we desire, subject only
to the restriction that we do not change the trace of $\obs^*\obs$, so
that $\sum_{i=1}^k \eta_i = m$. By concavity of $\log$, the highest
increase in the design criterion is gained by increasing
$\lambda_i^{-1} + \sigma^{-2}\eta_i$ where it is small. So, no weight
should be given to (i.e. measurement taken of) the $k$th eigenvector
before $\lambda_i^{-1} + \sigma^{-2}\eta_i \geq \lambda_k^{-1} \forall
i \neq k$. Thus, we should choose $\{\eta_i\}_{i=1}^k$ as follows. We
would increase $\lambda_1^{-1} + \sigma^{-2} \eta_1$ until it equals
$\lambda_2^{-1}$. Then, $\lambda_i^{-1} + \sigma^{-2} \eta_i,i=1,2$ is
increased until $\lambda_i^{-1} + \sigma^{-2} \eta_i =
\lambda_3^{-1},i=1,2$ and so forth.

@@

\subsection{Sensor Clusterization}\label{subsec:clusterization}
The symmetry mentioned in last paragraph of the previous section may
very well break in real life problems. If this happens and eigenvalues
of \eqref{eq:eigenproblem} are all different, then showing
sensors do cluster proceeds as follows. First, we show there
are only so many different equal, by the pigeonhole principle. In
order to make this argument work, one needs to show this eigenproblem
does, indeed, have only finitely many eigenvalues. This is done by
first recalling that $\fwd$ is strongly smoothing (so $\fwd \postcov
\fwd^*$ has fast decaying eigenvalues) and then showing that small
eigenvalues arise from un-informative measurements. We use the
following corollary, whose proof may be found in the appendix.
%
\begin{corollary}[Gain for No Model Error]\label{cor:zero mod err}
  If $\modcov = 0$, then
  \begin{equation*}
    \tar( \obs ) - \tar (\obsm )
    = -\frac12 \log (1 - \sigma^{-2} \langle \fwd \postcov \fwd^* \meas_m, \meas_m \rangle ).
  \end{equation*}
\end{corollary}
%
Suppose $\meas_j$ is a measurement with a small eigenvalue, i.e.\
\begin{equation}\label{eq:small eigenvalue}
  \sigma^{-2} \fwd \postcov \fwd^* \meas_j = \xi_j \meas_j,\ \xi_j \ll 1.
\end{equation}
By Corollary \ref{cor:zero mod err}, the increase in the design
criterion is
\begin{align*}
  \tar(\obs) - \tar(\obsm) &=
  -\frac12 \log (1- \sigma^{-2} \langle \fwd
  \postcovm \fwd^* \meas_j, \meas_j \rangle) \\
  %
  %
  %
  &= - \frac12 \log (1 - \xi_j) \text{ since $\|\meas_j\|=1$ }\\
  %
  %
  %
  &\approx \frac{\xi_j}{2}.
\end{align*}
Thus, negligibly small eigenvalues give negligible increase in the
design criterion and will not be a part of a truly optimal
design. Since the eigenvalues of $\fwd \postcov \fwd^*$ decay faster
than those of $\fwd\prcov\fwd^*$, the latter will determine how many
non-negligible eigenvalues $\fwd \postcov \fwd^*$ has. Thus, in the
eigenproblem \eqref{eq:eigenproblem}, we need only consider a finite
number of eigenvectors.

At this point, one would like to invoke the pigeonhole principle and
claim that eigenvectors cluster. This, however, is not as
straightforward as one may hope. If $\fwd \postcov \fwd^*$ only has
simple eigenvalues, then we may use the pigeonhole principle and
conclude that some eigenvectors are clustered (up to a sign). However,
if $\fwd \postcov \fwd^*$ has non-simple eigenvalues, one may hope to
somehow prove measurement vectors are either orthogonal or parallel,
which suffices for proving clusterization but that is not true.

\subsection{Non-Vanishing Model Error}
We can build on Lemma \ref{lemma:design increase} to find the gain in design
for identical measurements.
\begin{corollary}[Gain for Identical Measurement]\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{corollary}

In contrast to the previous case of vanishing model error, we will see
that if $\modcov \neq 0$ clustering will not occur. From Corollary
\ref{cor:same meas} we can observe that as $\sigma^2 \to 0$, the
increase in the design criterion gained by taking a measurement
identical to a previous one approaches zero. We may conclude that for
small measurement error levels, the clusterization effect will be
canceled by the presence of a non-zero model error. Since the design
criterion is not defined for $\sigma^2 = 0$ and identical
measurements, we cannot make a statement regarding $\sigma^2 = 0$,
except in the limiting sense described above.

\section{Conclusion}\label{section:conclusion}
Our relaxed model gives us insight to D-optimal designs. We see that
in an ideal settings, we should not reduce uncertainty in any
direction before it is the largest uncertainty present.

We believe the model presented is not too far from reality. After all,
taking any measurement (point evaluation or other) is just recording
the value of a linear combination of individual eigenvectors of the
prior. In light of this, we believe the reason for clusterization is close
to the argument made in section \ref{subsub:first attempt}. The reason
we observe clusterization in real life problems is probably due to
symmetry breaking, although we cannot show that rigorously.

@@

>>>>

\subsection{Characterizing D-Optimal Designs}\label{subsub:characterization}
In light of \eqref{eq:mod conditions}, let us denote eigenvalues of
$\fwd \prcov \fwd^*$ by $\lambda_1 \geq \lambda_2 \geq ...$ and its
eigenvectors $\{\ev_i\}_{i=1}^{\infty}$. The following lemma will be useful:

@@

Thus, if we take $\func_j^{*} = \meas_j$ and $C := (\fwd \prcov
\fwd^*)^{-1}$ in Lemma \ref{lemma:sim diag}, we conclude from the
eigenproblem \eqref{eq:mod conditions} that $\obs^*\obs$ has the same
eigenvectors as $\fwd \prcov \fwd^*$, namely
$\{\ev_i\}_{i=1}^{\infty}$ (infinitely many of them with zero
eigenvalue, since $\obs^*\obs$ is finite-rank). Denote the
corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$. By the argument
in the paragraph following \eqref{eq:mod conditions}:
\begin{align}\label{eq:pre manipulations}
  \begin{split}
    I + \sigma^{-2} \prcov^{1/2}  \fwd ^* \obs^* \obs \fwd \prcov^{1/2} &= 
    \prcov^{1/2} ( \prcov^{-1} + \sigma^{-2} \fwd^*\obs^*  \obs \fwd ) \prcov^{1/2}\\
    %
    %
    %
    &=\prcov^{1/2} \fwd^* ( \fwd^{-*} \prcov^{-1}\fwd^{-1} + \sigma^{-2} \obs^* \obs  ) \fwd \prcov^{1/2}\\
    %
    %
    %
    &= \prcov^{1/2} \fwd^* \left ( (\fwd \prcov \fwd^* )^{-1} +\sigma^{-2} \obs^* \obs  \right ) \fwd \prcov^{1/2}.
  \end{split}
\end{align}
By definition of the objective \eqref{eq:objective} and using
\eqref{eq:pre manipulations}:
\begin{align}\label{eq:mod objective}
  \begin{split}
    \tar ( \obs ) &=\frac12 \log \det \left( I + \sigma^{-2}
    \prcov^{1/2} \fwd ^* \obs^* \obs \fwd \prcov^{1/2} \right )\\
    %
    %
    %
    &=\frac12 \log \det
    \left ( (\fwd \prcov \fwd^* )^{-1} + \sigma^{-2}\obs^* \obs  \right ) \fwd \prcov \fwd^* \\ 
    %
    %
    %
    &=\frac12 \log \left (\prod_{i=1}^{\infty} ( \lambda_i^{-1} + \sigma^{-2} \eta_i ) \prod_{i=1}^{\infty} \lambda_i \right ) \\
    %
    %
    %
    &=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right ). 
  \end{split}
\end{align}
Therefore, an optimal design maximizes
\begin{equation}
  \sum_{i=1}^{k}\log (\lambda_i^{-1} + \sigma^{-2} \eta_i),
\end{equation}

\pgfplotstableread{
  Label          prior  $a_1$  $a_2$   $a_3$   $a_4$   $a_5$         $a_6$    topper
  $\ev_1$         0.2    0.8    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  $\ev_2$         0.8    0.2    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  $\ev_3$         2.2    0      0       0      0.2    0.33333333  0.3333333  0.001
  $\ev_4$         3.5    0      0       0      0      0           0          0.001
}\testdata

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\testdata}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=prior and posterior precision,
        xlabel=eigenvector,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=blue!60]   table [y=$a_1$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=red!60]    table [y=$a_2$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=black!60]  table [y=$a_3$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=orange!60] table [y=$a_4$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=cyan!60]   table [y=$a_5$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=purple!60] table [y=$a_6$, meta=Label, x expr=\coordindex] {\testdata};

      \addlegendentry{prior}
      \addlegendentry{$a_1$}
      \addlegendentry{$a_2$}
      \addlegendentry{$a_3$}
      \addlegendentry{$a_4$}
      \addlegendentry{$a_5$}
      \addlegendentry{$a_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Posterior precision per eigenvector. Each measurement
    $\func_i$ increases precision uniformly across the lowest precision
    eigenvectors. We see that $\func_2$ and $\func_3$ are repeated
    measurements, as well as $\func_5$ and $\func_6$.}
  \label{fig:clusterization}
\end{figure}
with the norm constraints on $\meas_j,j=1,...,m$. We show in the
appedix (Lemma \ref{lemma:free}) we can find $\meas_1,...,\meas_m$
such that $\{\eta_i\}_{i=1}^k$ have any value we desire, subject only
to the restriction that we do not change the trace of $\obs^*\obs$, so
that $\sum_{i=1}^k \eta_i = m$. By concavity of $\log$, the highest
increase in the design criterion is gained by increasing
$\lambda_i^{-1} + \sigma^{-2}\eta_i$ where it is small. So, no weight
should be given to (i.e. measurement taken of) the $k$th eigenvector
before $\lambda_i^{-1} + \sigma^{-2}\eta_i \geq \lambda_k^{-1} \forall
i \neq k$. Thus, we should choose $\{\eta_i\}_{i=1}^k$ as follows. We
would increase $\lambda_1^{-1} + \sigma^{-2} \eta_1$ until it equals
$\lambda_2^{-1}$. Then, $\lambda_i^{-1} + \sigma^{-2} \eta_i,i=1,2$ is
increased until $\lambda_i^{-1} + \sigma^{-2} \eta_i =
\lambda_3^{-1},i=1,2$ and so forth.

@@

\subsection{Sensor Clusterization}\label{subsec:clusterization}
The way clusterization can occur is best understood from figure
\ref{fig:clusterization}.

%% The symmetry mentioned in last paragraph of the previous section may
%% very well break in real life problems. If this happens and eigenvalues
%% of \eqref{eq:eigenproblem} are all different, then showing
%% sensors do cluster proceeds as follows. First, we show there
%% are only so many different equal, by the pigeonhole principle. In
%% order to make this argument work, one needs to show this eigenproblem
%% does, indeed, have only finitely many eigenvalues. This is done by
%% first recalling that $\fwd$ is strongly smoothing (so $\fwd \postcov
%% \fwd^*$ has fast decaying eigenvalues) and then showing that small
%% eigenvalues arise from un-informative measurements. We use the
%% following corollary, whose proof may be found in the appendix.
%% %
%% \begin{corollary}[Gain for No Model Error]\label{cor:zero mod err}
%%   If $\modcov = 0$, then
%%   \begin{equation*}
%%     \tar( \obs ) - \tar (\obsm )
%%     = -\frac12 \log (1 - \sigma^{-2} \langle \fwd \postcov \fwd^* \meas_m, \meas_m \rangle ).
%%   \end{equation*}
%% \end{corollary}
%% %
%% Suppose $\meas_j$ is a measurement with a small eigenvalue, i.e.\
%% \begin{equation}\label{eq:small eigenvalue}
%%   \sigma^{-2} \fwd \postcov \fwd^* \meas_j = \xi_j \meas_j,\ \xi_j \ll 1.
%% \end{equation}
%% By Corollary \ref{cor:zero mod err}, the increase in the design
%% criterion is
%% \begin{align*}
%%   \tar(\obs) - \tar(\obsm) &=
%%   -\frac12 \log (1- \sigma^{-2} \langle \fwd
%%   \postcovm \fwd^* \meas_j, \meas_j \rangle) \\
%%   %
%%   %
%%   %
%%   &= - \frac12 \log (1 - \xi_j) \text{ since $\|\meas_j\|=1$ }\\
%%   %
%%   %
%%   %
%%   &\approx \frac{\xi_j}{2}.
%% \end{align*}
%% Thus, negligibly small eigenvalues give negligible increase in the
%% design criterion and will not be a part of a truly optimal
%% design. Since the eigenvalues of $\fwd \postcov \fwd^*$ decay faster
%% than those of $\fwd\prcov\fwd^*$, the latter will determine how many
%% non-negligible eigenvalues $\fwd \postcov \fwd^*$ has. Thus, in the
%% eigenproblem \eqref{eq:eigenproblem}, we need only consider a finite
%% number of eigenvectors.

%% At this point, one would like to invoke the pigeonhole principle and
%% claim that eigenvectors cluster. This, however, is not as
%% straightforward as one may hope. If $\fwd \postcov \fwd^*$ only has
%% simple eigenvalues, then we may use the pigeonhole principle and
%% conclude that some eigenvectors are clustered (up to a sign). However,
%% if $\fwd \postcov \fwd^*$ has non-simple eigenvalues, one may hope to
%% somehow prove measurement vectors are either orthogonal or parallel,
%% which suffices for proving clusterization but that is not true.

\section{Analysis of Optimal Designs --- Non-Vanishing Model Error}
We can build on Lemma \ref{lemma:design increase} to find the gain in design
for identical measurements.
\begin{corollary}[Gain for Identical Measurement]\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{corollary}

In contrast to the previous case of vanishing model error, we will see
that if $\modcov \neq 0$ clustering will not occur. From Corollary
\ref{cor:same meas} we can observe that as $\sigma^2 \to 0$, the
increase in the design criterion gained by taking a measurement
identical to a previous one approaches zero. We may conclude that for
small measurement error levels, the clusterization effect will be
mitigated by the presence of a non-zero model error. Since the design
criterion is not defined for $\sigma^2 = 0$ and identical
measurements, we cannot make a statement regarding $\sigma^2 = 0$,
except in the limiting sense described above.

\section{Conclusion}\label{section:conclusion}
Our relaxed model gives us insight to D-optimal designs. We see that
in an ideal settings, we should not reduce uncertainty in any
direction (eigenvector) before it is the largest uncertainty present.

%% We believe the model presented is not too far from reality. After all,
%% taking any measurement (point evaluation or other) is just recording
%% the value of a linear combination of individual eigenvectors of the
%% prior. In light of this, we believe the reason for clusterization is close
%% to the argument made in section \ref{subsub:first attempt}. The reason
%% we observe clusterization in real life problems is probably due to
%% symmetry breaking, although we cannot show that rigorously.

@@

\section{Acknowledgements}
This study was conducted during when the author was a PhD student of
Prof. Georg Stadler in New York University's Courant Institute.
